ODGI: understanding pangenome graphs

### 2021-10-20 manuscript/main.tex
add when the figure is ready
shorter

The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium \citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
We anticipate that in the near future, studies will frequently begin from assemblies of similar quality, leading to demand for methods that allow us to create and explore pangenomes.
Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard workflows.

@@

Such a graph can be constructed by multiple sequence alignment \citep{Lee_2002,Grasso_2004}, but the resulting partially-ordered graph cannot represent complex variation like inversions and cycles that occur at structural scales in genomes.
General approaches to building pangenome graphs transitively reduce an alignment between sequences to an equivalent, labeled sequence graph \citep{Kehr_2014,Garrison_2019_thesis}.
Current methods to build these graphs are still in development \citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
Because ODGI make use of the GFA file standard its functionality can be mixed with other pangenome tools, such as the VG toolkit\citep{Garrison:2018}.
%The improvement of these methods depends on tools that allow us to investigate
This standardization supports the development of a reference set of tools that operate on the pangenome graph model.
%The need to understand the graphs made by these methods motivates it.
We require such a toolkit to understand the graphs made by these methods.
\FIXME{last sentence not really clear}

@@

\input{fig_odgi_viz}

>>>>

The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium \citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
We anticipate that assemblies of similar quality will become common, leading to demand for methods that allow us to create and explore pangenomes.
Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard workflows.

@@

Such a graph can be constructed by multiple sequence alignment \citep{Lee_2002,Grasso_2004}, or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph \citep{Kehr_2014,Garrison_2019_thesis}.
Current methods to build these graphs are still in development \citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
%\FIXME{may want to move GFA standard+VG to intro section 1}
%The improvement of these methods depends on tools that allow us to investigate
This standardization supports the development of a reference set of tools that operate on the pangenome graph model.
Such an effort began with the VG toolkit toolkit~\citep{Garrison:2018}.
Here we refocus it with with ODGI---a compatible, but independent set of algorithms focused on first-order interrogation and transformation of pangenome graphs.

%Because ODGI uses this standard to  can be mixed with other pangenome tools, such as the VG toolkit
%Here, we enhance ongoing efforts with the VG 
%The need to understand the graphs made by these methods motivates it.
%We require such a toolkit to understand the graphs made by these methods.
%\FIXME{last sentence not really clear}

@@

%\input{fig_odgi_viz}
